### 2024-05-29 journal/correctness.tex
working on comments from reviewer 1

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
If $B$ and $B'$ are in the same node, it follows from Lemma \ref{lem::headProgress} that their direct subblocks are disjoint.
Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

Now suppose $\var{v}$ is an internal node. By the definition of subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{createSumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}}
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

In this section, we show that the routines used to compute responses to \op{Dequeue}
operations work correctly.
We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  We do this by showing that the \fld{super} field of a block $B$ in node $\var{v}$ is read from  
\var{v}.\fld{parent}'s \var{head} field close to the time that $B$'s superblock $B_s$ is installed in $v$'s parent.
In particular, $B.super$ is written after $B$ is installed, but before $B_s$ is installed.
%On one hand, $B.super$ is written before $B_s$ is installed:
%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
%parent soon after.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ is not $\nl$.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

@@

\begin{lemma} \label{linearSat}
$L$ is a legal linearization ordering.
\end{lemma}
\begin{proof}
By Corollary \ref{cor::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
By Lemma \ref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
so it appears in $L$.
Also, if $op_{1}$ terminates before $op_{2}$ begins, then $op_{1}$ 
is propagated to the root before $op_2$ begins, so $op_1$ appears before $op_2$ in $L$.
\end{proof}

>>>>

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
If $B$ and $B'$ are in the same node, it follows from Lemma \ref{lem::headProgress} that their direct subblocks are disjoint.
Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

Now suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the enqueues in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these enqueues is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
\var{num\sub{enq}}
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of enqueues  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of dequeues contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

In this section, we show that the routines used to compute responses to \op{Dequeue}
operations work correctly.
We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  We do this by showing that \rev{the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.}
%On one hand, $B.super$ is written before $B_s$ is installed:
%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
%parent soon after.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of \rev{direct} subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ \rev{is assigned a non-\nl\ value before $B$'s superblock is installed}.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

@@

\begin{lemma} \label{linearSat}
$L$ is a legal linearization ordering.
\end{lemma}
\begin{proof}
By Corollary \ref{cor::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
By Lemma \ref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
so it appears in $L$.
\rev{Suppose an operation} $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
is propagated to the root before $op_2$ begins, \rev{and therefore before $op_2$ is propagated to the root.
Thus, $op_1$ appears in an earlier block of \var{root}.\fld{blocks} than $op_2$.
So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.}
\end{proof}
